Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-19 ms_preprint/ms_preprint.tex
added Wills corrections to the ms

William J. Rayment -- \textbf{do you have ORCID, Will?}

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions (Morrone, 2018; Vilhena \& Antonelli, 2015). Although there is often no agreement on the naming systems among proposed regionalization schema (but see Ebach et al., 2008), the hierarchical nature of bioregions is well recognized (Kreft \& Jetz, 2010; Vilhena \& Antonelli, 2015). Thus, bioregions are usually described within geographically larger classifications, such as `provinces' or `realms' (Costello et al., 2017; Spalding et al., 2007). The delimitation of bioregions is made through identifying sets of species that are endemic or have high levels of co-occurrence, and are distinct enough from other regions. In this sense, bioregions are interpreted as a natural representations of common historical and/or ecological processes shaped by physical and biological forces (Kreft \& Jetz, 2010). In particular, the identification of bioregions allows representation of the distribution of multiple species at a given time and space, and relating these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages (N. A. Hill et al., 2017; Lyons et al., 2017). Given these features, bioregions can be used as surrogates for biodiversity monitoring in data-poor regions and in regional conservation planning (Ferrier, 2002; Spalding et al., 2007), playing a fundamental role in conservation biogeography (Whittaker et al., 2005).

Seabird assemblages have been widely described and related to environmental characteristics (Hyrenbach et al., 2007; Ribic et al., 1997; Woehler et al., 2003). Considering the hierarchical nature of oceanographic processes and their resulting features (Haury et al., 1978), it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales (Haury et al., 1978; Hunt \& Schneider, 1987). In the marine environment, large scale (\textgreater100 km length scale) oceanographic features (e.g.~boundary currents and water masses) act as terrestrial biomes (based on vegetation), and influence seabird assemblage structure--i.e.~the occurrence of species and their relative numerical composition (Commins et al., 2014; Hyrenbach et al., 2007; Ribic et al., 1997). In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies (Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007). Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur (Davoren, 2013; Hazen, Suryan, et al., 2013). Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g.~water masses) to fine (e.g.~fronts) (Dunlop et al., 1988; Quillfeldt et al., 2015; Ribic et al., 1997). Despite extensive surveys (e.g. Ballance et al., 1997) and tracking studies (e.g. Carneiro et al., 2020), many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

One area in need of better understanding of seabird distributions is Australia, considered a global seabird hotspot (Karpouzi et al., 2007; Ramírez et al., 2017). Given the region's importance for seabird diversity, there are surprisingly few studies on seabird distributions at sea (Bernard et al., 2021; Mott \& Clarke, 2018). Off northwest Australia, at-sea surveys, multivariate analyses (Dunlop et al., 1988) and ensemble species distribution models of species assemblages (Lavers et al., 2014) showed that habitat use by tropical seabird species was related to sea surface temperatures and salinity. Dunlop et al. (1988) suggested that salinity was the main variable influencing species' distributions in oceanic waters while temperature was the key factor for species on the continental shelf. In contrast, there are few studies at large spatial and temporal scales off eastern Australia. T. A. Reid \& Hindell (2000) and Evans et al. (2021) analysed seabird assemblages and their relationships with oceanographic and prey data at coarse scales (tens of km) off southeast Tasmania. The remaining studies are mostly descriptive accounts locally (Blaber \& Milton, 1994; Norman, 1992), or focused on a single species (Priddel et al., 2014).

The East Australian Current (EAC) is the major western boundary current in the South Pacific Ocean, and the dominant boundary current off eastern Australia (Ganachaud et al., 2014). The EAC has been extensively studied, although research into the links between the EAC's oceanography and marine predators are not well known. The dynamic features associated with the EAC, such as mesoscale eddies and upwelling, are likely to influence marine biota (Scales, Miller, Hawkes, et al., 2014; Suthers et al., 2011). While most of the biological studies in this system were focused on plankton entrapment within the cores of eddies (Firme et al., 2023; Garcia et al., 2022; Hassler et al., 2011), there is a growing body of work focusing on fisheries (Brieva et al., 2015; Hobday, 2010; Young et al., 2011). These studies emphasized how the probability of occurrence of a given taxon, and the structure of species assemblages, were driven by the EAC. However, the relationships between oceanography and marine megafauna within the EAC are poorly known (Suthers et al., 2011). Some progress has been made using shark and ray movement data (Jaine et al., 2014; Niella et al., 2020; Niella et al., 2022), and satellite tracking has also demonstrated that increased sea surface temperatures in the EAC decreases little penguin \emph{Eudyptula minor} foraging success (Carroll et al., 2016), but information at the assemblage-level for seabirds in the EAC is absent.

Marine species range shifts are more likely to occur faster than terrestrial species due to their higher sensitivity to warming (Lenoir et al., 2020; Pinsky et al., 2019). As a result, climate change is rearranging the structure of species assemblages around the globe (Antão et al., 2020; Dornelas et al., 2014; Poloczanska et al., 2013). Top predators, for instance, may change their core habitats (Hazen, Jorgensen, et al., 2013; Reisinger et al., 2022), with unprecedented population consequences (Woehler \& Hobday, 2023). For seabirds, Péron et al. (2010) and Sojitra et al. (2022) showed that sea surface temperature anomalies affected the occurrence and abundances of albatrosses and petrels species with latitude, thus affecting assemblage structure. In eastern Australia, changing EAC conditions have affected a range of marine species. For example, changes in zooplankton assemblage structure and fish distribution patterns have been recorded and attributed to warming water temperatures, and shifts in shark distributions are expected (C. R. Johnson et al., 2011; Niella et al., 2022). A severe marine heatwave driven by a poleward strengthening of the EAC, showed drastic environmental impacts including disease outbreaks in aquaculture farms and wildlife mortalities (Oliver et al., 2017). Given that the EAC is strengthening and extending farther southward and getting warmer (Cai et al., 2005; Phillips et al., 2022), biological changes are expected to occur (Hobday, 2010; Poloczanska et al., 2007).

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We use the Region of Common Profiles {[}RCP; Foster et al. (2013){]}, a mixture model regarded as one of the best available tools for describing bioregions (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics (N. A. Hill et al., 2020; Warton et al., 2015). Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can compare to assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte, 2013). The South Caledonian Jet completes the EAC system at \textasciitilde22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (\textbf{Figure \ref{fig:fig1-study-area}a}; Kessler \& Cravatte (2013); Ganachaud et al. (2014)). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019; Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flow polewards and is primarily associated with the shelf break, but at \textasciitilde32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea (Mata et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension' (Oke, Pilo, et al. (2019); often termed the `Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the `EAC southern extension'; \textbf{Figure \ref{fig:fig1-study-area}a}), carrying eddy-like features as a transient current (Oke, Roughan, et al., 2019).

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods (Schreiber \& Burger, 2002). Therefore, seasonal differences in species composition were expected (e.g. K. Reid et al., 2023), and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude \(\times\) 1° longitude grid cells (c.~110 \(\times\) 110 km) and season (\textbf{Figure \ref{fig:fig1-study-area}b}), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \textgreater1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). Given a set of environmental variables, the model can predict assemblage probabilities of sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals (N. A. Hill et al., 2020; Warton et al., 2015).

The model is specified in terms of assemblages and site-specific probabilities. There are \(K\) assemblages (i.e.~RCPs) that are common across the study area, \(K = (1, ..., k)\). The outcome variable is a vector \(y_{i} = (y_{i1}, ... y_{iS})\), that give the count or presence-absence of each of the \(S\) species at site \(i\). Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site \(i\) has each of the \(K\) possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) (Foster et al., 2013). However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the \(K\) assemblage (Foster et al., 2013). As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as `sampling artifacts' (Foster et al., 2017), thus biasing results (Ponti \& Sannolo, 2023). In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e.~presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e.~averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the `ecomix' 1.0.0 package (S. Woolley et al., 2022), with the following workflow: (i) find the best number of groups (\(K\)) that describes the data, (ii) re-fit the model by fixing \(K\) and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (\textbf{Figure S3}). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (\textbf{Table \ref{tab:table2}}). The best models showed a good fit with no apparent deviations from the assumptions (\textbf{Figure S4}). All the best models selected `sst' or `sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its `sst' threshold between 18--21 °C, or `sss' 35.1 PSU for summer (\textbf{Figure \ref{fig:fig2-partial-plots}}, \textbf{Figure S5}), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected `bat', suggesting a gradient between `coastal' and `oceanic' taxa based on bathymetry (\textbf{Figure S5}). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the `climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \textasciitilde34°S, with both presence-absence and abundance models pointing to similar results (\textbf{Figure \ref{fig:fig3-point-pred}}). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (\textbf{Figure \ref{fig:fig4-prob-maps}}, \textbf{Figure S6}), showing that around \textasciitilde34°S, the `northern' RCP tended to have higher probabilities in coastal regions, whereas the `southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong `Tasmania/Bass Straight' assemblage appears (\textbf{Figure \ref{fig:fig3-point-pred}}, \textbf{Figure S6}).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g.~boobies \emph{Sula} spp., lesser frigatebird \emph{Fregata ariel}) and typical Southern Ocean species in the southern RCP (e.g.~small albatrosses \emph{Thalassarche} spp., fairy prion \emph{P. turtur}) (\textbf{Figure \ref{fig:fig5-spp-profiles}}, \textbf{Figure S7}). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the `northern' and `southern' RCPs. However, the intermediate `Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet \emph{Morus serrator} prominent with high probabilities of occurrence.

@@

The results do agree with the marine biogeographic units proposed by Costello et al. (2017), who identified a boundary at \textasciitilde34°S between their `Coral Sea' and `South Australia' realms. Their approach used 65,000 taxa, from benthic/pelagic and coastal/offshore habitats, thus integrating more complexity of the biogeographic boundaries in the marine environment. Other biological data also suggest `northern/southern' assemblages off eastern Australia, including tuna and billfish {[}also highly mobile top predators; Reygondeau et al. (2012){]}, mesopelagic fauna {[}which includes seabird prey; Sutton et al. (2017){]} and deep-sea invertebrates and fish (O'Hara, Williams, Althaus, et al., 2020; O'Hara, Williams, Woolley, et al., 2020). All these studies focused mainly on describing biological assemblages, as opposed to classifying environments. It is worth noting that the Longhurst Provinces approach was updated to include SST, SSS and bathymetry (the same variables retained by our best RCP models) and a temporal component to the classification (Reygondeau et al., 2013). The updated provinces are more similar to the seabird assemblages described here, although still not a perfect match, which could be partially explained because the seabird assemblage at EAC region includes several long-distance migratory species. Australia's bioregionalization has a long story of disagreements (Ebach, 2012), however, the inherent link that RCPs develop between environment and biological data can be a strong tool to support marine bioregionalization.

\hypertarget{fitting-rcp-models-to-high-mobile-animals}{%
\subsection{Fitting RCP models to high mobile animals}\label{fitting-rcp-models-to-high-mobile-animals}}

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales (Tremblay et al., 2009), they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well-established method to capture these relationships (Ainley et al., 2012; Hyrenbach et al., 2007; Woehler et al., 2003). Further, the spatial scales on which assemblages relate to their environment is variable and species-specific (Hunt \& Schneider, 1987; Ribic et al., 1997; Scales et al., 2017). Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude \(\times\) 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also \ref{sec:limitations}). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere (Harris et al., 2007; Santora \& Veit, 2013; Sojitra et al., 2022). Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to had been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.

>>>>

William J. Rayment -- \url{https://orcid.org/0000-0001-9558-5109}

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions (Morrone, 2018; Vilhena \& Antonelli, 2015). Although there is often no agreement on the naming systems among proposed regionalization schema (but see Ebach et al., 2008), the hierarchical nature of bioregions is well recognized (Kreft \& Jetz, 2010; Vilhena \& Antonelli, 2015). Thus, bioregions are usually described within geographically larger classifications, such as `provinces' or `realms' (Costello et al., 2017; Spalding et al., 2007). The delimitation of bioregions is made through identifying sets of species that are endemic or have high levels of co-occurrence, and are distinct enough from other regions. In this sense, bioregions are interpreted as natural representations of common historical and/or ecological processes shaped by physical and biological forces (Kreft \& Jetz, 2010). In particular, the identification of bioregions allows representation of the distribution of multiple species at a given time and space, and relating these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages (N. A. Hill et al., 2017; Lyons et al., 2017). Given these features, bioregions can be used as surrogates for biodiversity monitoring in data-poor regions and in regional conservation planning (Ferrier, 2002; Spalding et al., 2007), playing a fundamental role in conservation biogeography (Whittaker et al., 2005).

Seabird assemblages have been widely described and related to environmental characteristics (Hyrenbach et al., 2007; Ribic et al., 1997; Woehler et al., 2003). Considering the hierarchical nature of oceanographic processes and their resulting features (Haury et al., 1978), it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales (Haury et al., 1978; Hunt \& Schneider, 1987). In the marine environment, large scale (\textgreater100 km length) oceanographic features (e.g.~boundary currents and water masses) act as terrestrial biomes (based on vegetation), and influence seabird assemblage structure--i.e.~the occurrence of species and their relative numerical composition (Commins et al., 2014; Hyrenbach et al., 2007; Ribic et al., 1997). In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies (Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007). Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur (Davoren, 2013; Hazen, Suryan, et al., 2013). Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g.~water masses) to fine (e.g.~fronts) (Dunlop et al., 1988; Quillfeldt et al., 2015; Ribic et al., 1997). Despite extensive surveys (e.g. Ballance et al., 1997) and tracking studies (e.g. Carneiro et al., 2020), many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

One area in need of better understanding of seabird distributions is Australia, considered a global seabird hotspot (Karpouzi et al., 2007; Ramírez et al., 2017). Given the region's importance for seabird diversity, there are surprisingly few studies on seabird distributions at sea (Bernard et al., 2021; Mott \& Clarke, 2018). Off northwest Australia, at-sea surveys, multivariate analyses (Dunlop et al., 1988) and ensemble species distribution models of species assemblages (Lavers et al., 2014) showed that habitat use by tropical seabird species was related to sea surface temperatures and salinity. Dunlop et al. (1988) suggested that salinity was the main variable influencing species' distributions in oceanic waters while temperature was the key factor for species on the continental shelf. In contrast, there are few studies at large spatial and temporal scales off eastern Australia. T. A. Reid \& Hindell (2000) and Evans et al. (2021) analysed seabird assemblages and their relationships with oceanographic and prey data at mesoscale off southeast Tasmania. The remaining studies are mostly descriptive accounts locally (Blaber \& Milton, 1994; Norman, 1992), or focused on a single species (Priddel et al., 2014).

The East Australian Current (EAC) is the major western boundary current in the South Pacific Ocean, and the dominant boundary current off eastern Australia (Ganachaud et al., 2014). The EAC has been extensively studied, although the links between the EAC's oceanography and marine predators are not well known. The dynamic features associated with the EAC, such as mesoscale eddies and upwelling, are likely to influence marine biota (Scales, Miller, Hawkes, et al., 2014; Suthers et al., 2011). While most of the biological studies in this system were focused on plankton entrapment within the cores of eddies (Firme et al., 2023; Garcia et al., 2022; Hassler et al., 2011), there is a growing body of work focusing on fisheries (Brieva et al., 2015; Hobday, 2010; Young et al., 2011). These studies emphasized how the probability of occurrence of a given taxon, and the structure of species assemblages, were driven by the EAC. However, the relationships between oceanography and marine megafauna within the EAC are poorly known (Suthers et al., 2011). Some progress has been made using shark and ray movement data (Jaine et al., 2014; Niella et al., 2020; Niella et al., 2022), and satellite tracking has also demonstrated that increased sea surface temperatures in the EAC decreases little penguin \emph{Eudyptula minor} foraging success (Carroll et al., 2016), but information at the assemblage-level for seabirds in the EAC is absent.

Marine species range shifts are more likely to occur faster than terrestrial species due to their higher sensitivity to warming (Lenoir et al., 2020; Pinsky et al., 2019). As a result, climate change is rearranging the structure of species assemblages around the globe (Antão et al., 2020; Dornelas et al., 2014; Poloczanska et al., 2013). Top predators, for instance, may change their core habitats (Hazen, Jorgensen, et al., 2013; Reisinger et al., 2022), with unprecedented population consequences (Woehler \& Hobday, 2023). For seabirds, Péron et al. (2010) and Sojitra et al. (2022) showed that sea surface temperature anomalies affected the occurrence and abundances of albatross and petrel species with latitude, thus affecting assemblage structure. In eastern Australia, changing EAC conditions have affected a range of marine species. For example, changes in zooplankton assemblage structure and fish distribution patterns have been recorded and attributed to warming water temperatures, and shifts in shark distributions are expected (C. R. Johnson et al., 2011; Niella et al., 2022). A severe marine heatwave driven by a poleward strengthening of the EAC, showed drastic environmental impacts including disease outbreaks in aquaculture farms and wildlife mortalities (Oliver et al., 2017). Given that the EAC is strengthening and extending farther southward and getting warmer (Cai et al., 2005; Phillips et al., 2022), biological changes are expected to occur (Hobday, 2010; Poloczanska et al., 2007).

@@

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea survey data and a state-of-the-art modelling technique. We use the Region of Common Profiles {[}RCP; Foster et al. (2013){]}, a mixture model regarded as one of the best available tools for describing bioregions (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics (N. A. Hill et al., 2020; Warton et al., 2015). Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate. We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte, 2013). The South Caledonian Jet completes the EAC system at \textasciitilde22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (\textbf{Figure \ref{fig:fig1-study-area}a}; Kessler \& Cravatte (2013); Ganachaud et al. (2014)). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019; Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flows polewards and is primarily associated with the shelf break, but at \textasciitilde32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea (Mata et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension' (Oke, Pilo, et al. (2019); often termed the `Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the `EAC southern extension'; \textbf{Figure \ref{fig:fig1-study-area}a}), carrying eddy-like features as a transient current (Oke, Roughan, et al., 2019).

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods (Schreiber \& Burger, 2002). Therefore, seasonal differences in species composition were expected (e.g. K. Reid et al., 2023), and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude \(\times\) 1° longitude grid cells (\emph{c.} 110 \(\times\) 110 km) and season (\textbf{Figure \ref{fig:fig1-study-area}b}), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \textgreater1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) (N. A. Hill et al., 2020; S. N. C. Woolley et al., 2020). Given a set of environmental variables, the model can predict assemblage probabilities for sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals (N. A. Hill et al., 2020; Warton et al., 2015).

The model is specified in terms of assemblages and site-specific probabilities. There are \(K\) assemblages (i.e.~RCPs), \(K = (1, ..., k)\), that are common across the study area. The outcome variable is a vector \(y_{i} = (y_{i1}, ... y_{iS})\), that gives the count or presence-absence of each of the \(S\) species at site \(i\). Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site \(i\) has each of the \(K\) possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) (Foster et al., 2013). However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the \(K\) assemblages (Foster et al., 2013). As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

@@

Due to seabird migration and dispersion, seasons could act as `sampling artifacts' (Foster et al., 2017), thus biasing results (Ponti \& Sannolo, 2023). In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \ref{sec:modellingdatasets}), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e.~presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e.~averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data were based on a Negative Binomial model. RCPs were fitted using the `ecomix' 1.0.0 package (S. Woolley et al., 2022), with the following workflow: (i) find the best number of groups (\(K\)) that describes the data, (ii) re-fit the model by fixing \(K\) and selecting the optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (\textbf{Figure S3}). The best fitting models generally had one or two covariates selected, although the abundance-autumn model had five covariates (\textbf{Table \ref{tab:table2}}). The best models showed a good fit with no apparent deviations from the assumptions (\textbf{Figure S4}). All the best models selected `sst' or `sss', as essential for characterising water masses. The thresholds for defining RCP membership were 18--21°C for `sst' and 35.1 PSU for `sss' (\textbf{Figure \ref{fig:fig2-partial-plots}}, \textbf{Figure S5}), similar to the characteristics used to to define the EAC. Furthermore, most of the best models selected `bat', suggesting a gradient between `coastal' and `oceanic' taxa based on bathymetry (\textbf{Figure S5}). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the `climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \textasciitilde34°S, with both presence-absence and abundance models pointing to similar results (\textbf{Figure \ref{fig:fig3-point-pred}}). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (\textbf{Figure \ref{fig:fig4-prob-maps}}, \textbf{Figure S6}), showing that around \textasciitilde34°S, the `northern' RCP tended to have higher probabilities in coastal regions, whereas the `southern' RCP spread northward offshore. The autumn presence-absence model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong `Tasmania/Bass Strait' assemblage appears (\textbf{Figure \ref{fig:fig3-point-pred}}, \textbf{Figure S6}).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g.~boobies \emph{Sula} spp., lesser frigatebird \emph{Fregata ariel}) and typical Southern Ocean species in the southern RCP (e.g.~small albatrosses \emph{Thalassarche} spp., fairy prion \emph{P. turtur}) (\textbf{Figure \ref{fig:fig5-spp-profiles}}, \textbf{Figure S7}). The autumn presence-absence model, which had three RCPs, also showed the characteristic taxa representing the `northern' and `southern' RCPs. However, the intermediate `Tasmania/Bass Strait' assemblage showed a mixture of both groups with shy albatross and Australasian gannet \emph{Morus serrator} prominent with high probabilities of occurrence.

@@

The results do agree with the marine biogeographic units proposed by Costello et al. (2017), who identified a boundary at \textasciitilde34°S between their `Coral Sea' and `South Australia' realms. Their approach used 65,000 taxa, from benthic/pelagic and coastal/offshore habitats, thus integrating more complexity of the biogeographic boundaries in the marine environment. Other biological data also suggest `northern/southern' assemblages off eastern Australia, including tuna and billfish {[}also highly mobile top predators; Reygondeau et al. (2012){]}, mesopelagic fauna {[}which includes seabird prey; Sutton et al. (2017){]} and deep-sea invertebrates and fish (O'Hara, Williams, Althaus, et al., 2020; O'Hara, Williams, Woolley, et al., 2020). All these studies focused mainly on describing biological assemblages, as opposed to classifying environments. It is worth noting that the Longhurst Provinces approach was updated to include SST, SSS and bathymetry (the same variables retained by our best RCP models) and a temporal component to the classification (Reygondeau et al., 2013). The updated provinces are more similar to the seabird assemblages described here, although still not a perfect match, which could be partially explained because the seabird assemblage in the EAC region includes several long-distance migratory species. Australia's bioregionalization has a long story of disagreements (Ebach, 2012), however, the inherent link that RCPs develop between environment and biological data can be a strong tool to support marine bioregionalization.

\hypertarget{fitting-rcp-models-to-highly-mobile-animals}{%
\subsection{Fitting RCP models to highly mobile animals}\label{fitting-rcp-models-to-highly-mobile-animals}}

@@

While remote tracking studies are advancing our understanding on how individual seabirds use their environment at fine scales (Tremblay et al., 2009), they fail to perceive how species assemblages interact with the environment and each other. Therefore, using at-sea counts of seabirds is a well established method to capture these relationships (Ainley et al., 2012; Hyrenbach et al., 2007; Woehler et al., 2003). Further, the spatial scales on which assemblages relate to their environment is variable and species-specific (Hunt \& Schneider, 1987; Ribic et al., 1997; Scales et al., 2017). Understanding the biology of the studied group is thus essential to set the appropriate scales of analyses. Our choice of 1° latitude \(\times\) 1° longitude was mainly determined by data availability reflecting survey effort, and the trade-off between the number of species per grid/season (see also \ref{sec:limitations}). Nonetheless, the chosen spatial scale is similar to other seabird distribution analyses conducted elsewhere (Harris et al., 2007; Santora \& Veit, 2013; Sojitra et al., 2022). Therefore, for describing seabird assemblages at a macro-scale, the chosen resolution seems to have been a good fit, accords with studies elsewhere, and the resultant models described patterns of seabird species' composition consistent with our current understanding of species' ecologies and life history strategies.
